---
output:
  html_document: default
  pdf_document: default
---

1.

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')


setwd("C:\\Users\\T430\\Documents\\IODS-project-1\\data")

RATSL= read.table("RATSL.txt")


str(RATSL)

RATSL$Group <-factor(RATSL$Group)
RATSL$ID <- factor(RATSL$ID)

BPRSL <- read.table("BPRSL.txt", header = TRUE)
glimpse(BPRSL)

head(RATSL)
```

```{r}


library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = ((Weight - mean(Weight))/sd(Weight))) %>%
  ungroup()
```

# Plot again with the standardised bprs
```{r}
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")
```

```{r}
n <- RATSL$Time %>% unique() %>% length()


# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
```


```{r}
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")


```

```{r}
RATSL64S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(average=mean(Weight)) %>%
  ungroup()
# Glimpse the data
glimpse(RATSL64S)
```



```{r}
# Draw a boxplot of the mean versus treatment
ggplot(RATSL64S, aes(x = Group, y = average)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), periods 8-64")
```


```{r}
RATSL64SF <- RATSL64S %>% filter(average < 580, average > 240, average != 495.2)
ggplot(RATSL64SF, aes(x = Group, y = average)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), periods 8-64")
```

```{r}


RATSL8S2 <- RATSL64S %>%
  mutate(baseline = RATS$WD1)

```


```{r}
fit <- lm(average ~ baseline + Group, data = RATSL8S2)
# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```




The data has 360 observations and 5 variables. The variables describe the sum of the scale measures levels, what treatment group the individual is assigned, and when the measures were collected.


```{r}
summary(BPRSL$bprs)
summary(BPRSL$treatment)
BPRSL$treatment <- factor(BPRSL$treatment)
```

The median of bprs rating is 35 and mean is 37.66. Some individuals diviate from this average signficantly, since the minimum of the rating is 18 and maximum 95.

```{r}
BPRSL$subject <- factor(BPRSL$subject)
BPRSL$group <- factor(BPRSL$treatment)
ggplot(BPRSL, aes(x = week, y= bprs, linetype = subject, col = treatment)) +
  geom_line() +
   scale_linetype_manual(values = rep(1:10, times=4)) +
 
    theme(legend.position = "none") +
    scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

In the blue are people who received treatment 2, and in red people received treatment 1. Maybe we get a clearer picture, if we plot these two groups individually by group:

```{r}
BPRSL$subject <- factor(BPRSL$subject)
BPRSL$group <- factor(BPRSL$treatment)
ggplot(BPRSL, aes(x = week, y= bprs, linetype = subject, col = treatment)) +
  geom_line() +
   scale_linetype_manual(values = rep(1:10, times=4)) +
      facet_grid(. ~ treatment, labeller = label_both) +
      theme(legend.position = "none") +
    scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```


```{r}
Fitted <- fitted(RATS_ref2)

# Create a new column fitted to RATSL
RATSL = mutate(RATSL,Fitted)
```

```{r}
ggplot(BPRSL, aes(x = week, y =Fitted, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")
```

It may be difficult to distinguish a clear trend from these plots. For some people, it seems that treatment in group 1 or group 2 decreased BRPS points, but for some the treatment period did not.

```{r}
# create a regression model
BPRSL_reg <- lm(bprs ~ week + treatment, data = BPRSL)
# print out a summary of the model
summary(BPRSL_reg)
```

From a simple regression model bprs = week + treatment, we see that the p-value for treatment2 is 0.661. Therefore, we cannot reject the null hypothesis that the coefficient of treatment2 is zero. Here we assume unrealistically indepence to the repeated measures of bprs for individuals.

```{r}
# access library lme4
library(lme4)
# Fitting a random intercept model allows the linear regression fit for each rat to differ # in intercept from other rats.
# Create a random intercept model
BPRSL_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
# Print the summary of the model
summary(BPRSL_ref)
```

With the above random intercept model, we can allow individual BPRS to differ with an intercept term, that is usually not the same as for other individuals. Here, the standard errors are lower than in the simple linear regression model. The conclusion does not change in this model.

```{r}
# create a random intercept and random slope model
BPRSL_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
# Fitting a random intercept and random slope model allows the linear regression fits for each individual to differ in intercept but also in slope. This way it is possible to account for the individual differences in the rats' growth profiles, but also the   effect of time.
# print a summary of the model
summary(BPRSL_ref1)
# perform an ANOVA test on the two models
anova(BPRSL_ref1, BPRSL_ref)
```

Fitting the random intercept and slope model to the BPRS data results compared to the
the fixed effects model are very similar. However, the chi-squared statistics and p-value of the likelihood ratio test between BPRSL_ref and BPRS_ref1 are very low, indicating that the fit is better in BPRS_ref1, that is compared to BPRS_ref model.

```{r}
# create a random intercept and random slope model
# Add a interaction term
BPRSL_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
# print a summary of the model
summary(BPRSL_ref2)
# perform an ANOVA test on the two models
anova(BPRSL_ref2, BPRSL_ref1)
```

Here, I created a random intercept and random slope model with an interaction term of week and treatment, measuring how these terms affect together bprs value.

The chi-squared statistics is low (3.17). P-value of the likelihood ratio test between BPRSL_ref1 and BPRS_ref2 is 0.07495, indicating that on 10 % significance level, the model increases fit. However, on 5 % significance level, we cannot make this conclusion.

```{r}
# Create a vector of the fitted values
Fitted <- fitted(BPRSL_ref2)
# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>%
  mutate(Fitted)
ggplot(BPRSL, aes(x = week, y= Fitted, linetype = subject, col = treatment)) +
  geom_line() +
   scale_linetype_manual(values = rep(1:10, times=4)) +
      facet_grid(. ~ treatment, labeller = label_both) +
      theme(legend.position = "none") +
    scale_y_continuous(limits = c(min(BPRSL$Fitted), max(BPRSL$Fitted)))
```
